R Notebook for R Review - ANSWER KEY
QBIO Public Data Analysis Student Group
3/12/2021

Here is a review R Notebook. This is designed to be useful for the small group project. 
```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
if(!requireNamespace("devtools")) BiocManager::install(c("devtools"))
if(!requireNamespace("robustbase"))BiocManager::install(c("robustbase"))
library(devtools)
library(robustbase)

devtools::install_github("BioinformaticsFMRP/TCGAbiolinks")

if(!requireNamespace("SummarizedExperiment"))BiocManager::install(c("SummarizedExperiment"))
if(!requireNamespace("maftools"))BiocManager::install(c("maftools"))
if(!requireNamespace("arsenal"))install.packages(c("arsenal"))
if(!requireNamespace("survival"))install.packages(c("survival"))
if(!requireNamespace("survminer"))install.packages(c("survminer"))

library(TCGAbiolinks)
library(maftools)
library(SummarizedExperiment)
library(arsenal)
library(survival)
library(survminer)
```

Use these barcodes if using on your local machine as an example for smaller files downloaded
```{r}
barcodes_rnaseq <- c("TCGA-BH-A0DG-01A-21R-A12P-07","TCGA-A2-A0YF-01A-21R-A109-07",
          "TCGA-AN-A04A-01A-21R-A034-07","TCGA-AR-A1AP-01A-11R-A12P-07",
           "TCGA-A2-A0T3-01A-21R-A115-07", "TCGA-E2-A154-01A-11R-A115-07" )
barcodes_clinic <- c("TCGA-BH-A0DG","TCGA-A2-A0YF","TCGA-AN-A04A","TCGA-AR-A1AP", "TCGA-A2-A0T3",
                      "TCGA-E2-A154", "TCGA-AO-A12F", "TCGA-A2-A0YM", "TCGA-BH-A0E0", "TCGA-AN-A0FL")
```

PART 1: DATAFRAMES

Load in the subtype data
1. Use head() to look at the subtypes dataframe
  a. What columns are in subtypes?
  b. Which columns do you personally find interesting and why?
```{r}
subtypes <- TCGAquery_subtype(tumor = "BRCA")
head(subtypes)
colnames(subtypes)
```
2. Use str() and dim() to look at subtypes
  a. How is str() different than head?
  b. How many rows and columns are in subtypes?
  c. How many patients are in subtypes?
```{r}
str( subtypes )
dim( subtypes )
```

3. Standard Indexing
  a. Use the format, dataframe[row, col] to find the element at row 4 and col 3.
  b. Removing the row or col in [row,col] allows you to access an entire row or column. Access all columns of the 7th row. 
  c. 1:10 is a special R way to create a list of from 1 to 10. Type 1:10 in the console if this is unclear. 
     Access all columns of the first 10 rows. Note this is the first 10 patients.
  d. Is the format subtypes[row,col] useful for this dataframe? Why or why not?
  e. Don't forget about the $. Create a variable called BRCA_PAM50 and assign the BRCA_Subtype_PAM50 column to it. 
```{r}
subtypes[1:10,]
BRCA_PAM50 <- subtypes$BRCA_Subtype_PAM50
```

4. Boolean Indexing
Boolean Indexing is a unique and helpful way to index into a dataframe.
It utilizes a "boolean mask". A boolean is a TRUE or FALSE element. A boolean mask is a list of TRUE's and FALSE's.
Follow the following steps to find patients less than 40yrs old AND are in the LumA subtype
  1. Describe the contents of luma_mask. Note that "==" means "equal to"
  2. Write an equivalent line of code that creates the variable luma_mask_check that is equivalent to luma_mask,
     but uses the BRCA_PAM50 variable above.
  3. Computers read TRUE as "1" and FALSE as "0". A cool trick is to sum the elements of the mask to see how many TRUE elements exist.
     Using sum() and as.numeric(luma_mask), how many patients are Luminal A subtype?
  4. Use information from ^^ check that luma_mask and luma_mask_check have the same number of TRUE elements.
  5. Create a boolean mask called young_mask that is TRUE when the patient is less than (<) 40 yrs old. How many patients are under 40?
     How many patients are over 40 in subtype?
  6. Create a boolean mask called young_and_luma_mask that is TRUE when the patient in LumA and young. Use the "&" symbol.
  7. You can use the boolean mask to index into the subtypes dataframe. Remember part 3. Standard Indexing. 
     Create a new dataframe called young_and_luma_dataframe and assign it to all columns of the patients that are young and luminal a
     using your young_and_luma_mask. Hint: This is the same format as the 3c using the 1:10. 
  8. What if we just want which patients are young and luminal a? You can also index into a specific column. 
     Using $ and subtypes[young_and_luma_mask], create a variable young_and_luma_patients with the list of young and luminal a patients.
```{r}
luma_mask <- subtypes$BRCA_Subtype_PAM50 == "LumA"
luma_mask_check <- BRCA_PAM50 == "LumA"
sum(as.numeric(luma_mask)) == sum(as.numeric(luma_mask_check))
young_mask <- subtypes$age_at_initial_pathologic_diagnosis < 40
sum(as.numeric(young_mask))
young_and_luma_mask <- luma_mask & young_mask
young_and_luma_dataframe <- subtypes[ young_and_luma_mask, ]
young_and_luma_patients <- subtypes$patient[ young_and_luma_mask ]
```

4.5. Use masks to summarize data.
  a. What percentage of the young (<40) patients in the subtypes data frame are Luminal A?
  b. What percentage of the old (>=60) patients in the subtypes data frame are Luminal A?
```{r}

```


PART 2: RNASeq data

Download and/or access data
```{r}
query <- GDCquery(project = "TCGA-BRCA",
                   data.category = "Transcriptome Profiling",
                   data.type = "Gene Expression Quantification",
                   workflow.type = "HTSeq - Counts",
                   barcode = barcodes_rnaseq)
GDCdownload(query) #only need this line of code ONCE to download the data
sum_exp <- GDCprepare(query)
```

1. SummarizedExperiment 
   a. What does sum_exp include?
   b. What does colData(sum_exp) include?
   c. What does rowData(sum_exp) include?
   d. What does assays(sum_exp) include?
   e. Describe assays(sum_exp)$"HTSeq - Counts"
```{r}
colData(sum_exp)$paper_age_at_diagnosis
colnames(colData(sum_exp))
```

2. ifelse
Remember the ifelse structure:
ifelse(condition, action when condition true, action when condition false)
   a.Create a new column in colData(sum_exp) called age_category with the following age categories:
     < 40 is "young", 40-59 is "mid", 60+ is "old".
     Use colData(sum_exp)
```{r}
colData(sum_exp)$age_category <- ifelse(colData(sum_exp)$paper_age_at_initial_pathologic_diagnosis < 40, "Young", ifelse(colData(sum_exp)$paper_age_at_initial_pathologic_diagnosis >= 60, "Old", "Mid"))
```

3. Applying boolean masks to HTSeq
   a. Create variable young_htseq_mask using colData(sum_exp)$age_category
   b. Create variable htseq_data that consists of: assays(sum_exp)$"HTSeq - Counts"
   c. Create variable young_htseq_data that consists of all genes (rows) of only the young patients (cols). Run str(young_htseq_data).
   d. Use dim(young_htseq_data)[2] == sum(as.numeric()) to check that your code is correct.
   e. Why does (d) verify if your code is correct?
```{r}
young_htseq_mask <- colData(sum_exp)$age_category == "Young"
htseq_data <- assays(sum_exp)$"HTSeq - Counts"
young_htseq_data <- htseq_data[, young_htseq_mask]
dim(young_htseq_data)[2] == sum(as.numeric(young_htseq_mask))
```

4. Note about barcodes
Notice that in colnames(colData(sum_exp)) there are two columns that discuss barcodes: 
                                                    colData(sum_exp)$barcode and colData(sum_exp)$bcr_patient_barcode
Notice the barcode is longer with extra numbers and letters. This is because it refers the the tumor SAMPLE.
The bcr_patient_barcode refers to the PATIENT. You can imagine that one PATIENT can have multiple tumor SAMPLES.
The bcr_patient_barcode is the same in the clinical data, and the subtypes dataframe above. 
In your project, you can use this barcode to interface between breast cancer subtype (an important factor in breast cancer) and the RNASeq counts.
```{r}
head( colData(sum_exp)$barcode )
head( colData(sum_exp)$bcr_patient_barcode )
```


